---
title: "GSE6791_HNSCC_HPV_Tumor"
author: "Shruti"
date: "November 20, 2015"
output: html_document
---
**HPV positive vs HPV negative analysis for Head and Neck cancer dataset - GSE6791**  

**Tumor vs Normal analysis for Head and Neck cancer dataset - GSE6791 under different conditions:** 
** Tumor (HPV positive & negative) vs Normal (HPVnegative); 
HPV Positive Tumor vs Normal (HPV negative); 
HPV Negative Tumor vs Normal (HPVnegative); 
Tumor vs Normal - tonsil samples only **

**Load Required Packages**
```{r Chunk 1: load packages}
# if a new package from bioconductor has to be installed. write following two commands
# source("https://bioconductor.org/biocLite.R")
# biocLite("package name")
library(GEOquery)
library(limma)
#install.packages("rafalib")
library("rafalib")
```

**Download input data**
```{r Chunk 2: download data}
gset6791 <- getGEO("GSE6791", GSEMatrix =TRUE)
#setwd("/Users/shruti/Dropbox/SHRUTIM/Microarray/Microarray_R_Scripts/HNSCC/TumorVSControl/GSE6791/")
#save(gset6791,file="./final/gset6791.rda")
#load("/Users/shruti/Dropbox/SHRUTIM/Microarray/Microarray_R_Scripts/HNSCC/TumorVSControl/GSE6791/final/gset6791.rda")
```

**Data Munging**
```{r Chunk 3: data munging}
GPLid <- levels((gset6791[[1]])$platform_id)
if (length(gset6791) > 1) 
{
  idx <- grep(GPLid, attr(gset6791, "names"))
} else 
{ 
  idx <- 1
}
gset6791 <- gset6791[[idx]]

## look at the pattern in which the sample types are arranged. 
#colnames(pData(gset6791))
#pData(gset6791)[1:2,]
#pData(gset6791)[1:2,c(1,7,10:14,19,24,25)]
#apply(pData(gset6791)[,c(10:12,14,24)],2,table)

# since HPV information is not available through gset, downlaod its meta data
dfl6791 <- getGSEDataTables("GSE6791")
#save(dfl6791,file="./final/dfl6791.rda")
#load("./final/dfl6791.rda")
sample_info <- dfl6791[c(1:12),]
sample_info <- as.data.frame(sample_info,stringsAsFactors=F)
sample_info <- as.data.frame(apply(sample_info,2,function(i) {sub("\\s$","",i)}))
rownames(sample_info) <- sample_info$Accession.number
colnames(sample_info) <- sapply(colnames(sample_info),function(i) {sub("\\.$","",i)},USE.NAMES=F)

# select on samples for Head and Neck tissue samples
#pData(gset6791)[1:2,1:2]
#sample_info[1:2,]
HNnormal_gsmids <-  rownames(subset(sample_info,Case=="HN normal",))
HNcancer_gsmids <-  rownames(subset(sample_info,Case=="HN cancer",))

gset6791_HN_tissue <- gset6791[,c(HNnormal_gsmids,HNcancer_gsmids)]
# log2 transform
ex6791_HN_tissue <- exprs(gset6791_HN_tissue)
qx <- as.numeric(quantile(ex6791_HN_tissue, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
LogC <- (qx[5] > 100) ||
  (qx[6]-qx[1] > 50 && qx[2] > 0) ||
  (qx[2] > 0 && qx[2] < 1 && qx[4] > 1 && qx[4] < 2)
if (LogC) { ex6791_HN_tissue[which(ex6791_HN_tissue <= 0)] <- NaN
            exprs(gset6791_HN_tissue) <- log2(ex6791_HN_tissue) }

sample_info_HN_tissue <- sample_info[c(HNnormal_gsmids,HNcancer_gsmids),]
#apply(sample_info_HN_tissue[,c(2:8)],2,table)

sample_info_HN_tissue$HPV <- sub(pattern="\\s?[(]\\d+[)]", replacement="", x=sample_info_HN_tissue$HPV)

cancer_status <- sample_info_HN_tissue$Case
cancer_status <- sub(pattern="HN ", replacement="", x=cancer_status,ignore.case=T)
cancer_status_factors <-  factor(cancer_status, levels=c("normal","cancer"))

HPV_status_factors <- factor(sample_info_HN_tissue$HPV, levels=c("-","+"))
cancer_HPV_status <- factor(paste(cancer_status_factors,HPV_status_factors,sep=""),levels=c("normal-","cancer-","cancer+"))

sex <- factor(sample_info_HN_tissue$Gender)
tissue <- factor(sample_info_HN_tissue$Anatomical.sites)
cancer_tissue_status <- factor(paste(cancer_status_factors,tissue,sep=" "))
```

**Exploratory Data Analysis for Tumor vs Normal**
```{r Chunk 4:  Exploratory Data Analysis for Tumor vs Normal}
dist_matrix_HN_tissue_6791 <- dist(t(ex6791_HN_tissue)) # very important to take transpose "t"
mds_HN_tissue_6791 <- cmdscale(dist_matrix_HN_tissue_6791)

mypar(1,1)
# cancer_status
plot(mds_HN_tissue_6791[,1],mds_HN_tissue_6791[,2],bg=as.numeric(cancer_status_factors),pch=21,xlab="First dimension",ylab="Second dimension",cex=2,main="MultiDimension Scaling Plot: cancer status")
legend("bottomleft",levels(cancer_status_factors),col=seq(along=levels(cancer_status_factors)),pch=15,cex=1)
text(mds_HN_tissue_6791[,1],mds_HN_tissue_6791[,2],labels=rownames(mds_HN_tissue_6791),cex=0.6)
#identify(mds_HN_tissue_6791)
sample_info_HN_tissue[c(4,55,2,19),c(2,4,6)]

# cancer and HPV (+ or -)status
plot(mds_HN_tissue_6791[,1],mds_HN_tissue_6791[,2],bg=as.numeric(cancer_HPV_status),pch=21,xlab="First dimension",ylab="Second dimension",cex=2,main="MultiDimension Scaling Plot: cancer & HPV status")
legend("bottomleft",levels(cancer_HPV_status),col=seq(along=levels(cancer_HPV_status)),pch=15,cex=1)

# HPV (+ or -)status
plot(mds_HN_tissue_6791[,1],mds_HN_tissue_6791[,2],bg=as.numeric(HPV_status_factors),pch=21,xlab="First dimension",ylab="Second dimension",cex=2,main="MultiDimension Scaling Plot: HPV status")
legend("bottomleft",levels(HPV_status_factors),col=seq(along=levels(HPV_status_factors)),pch=15,cex=1)

# sex
plot(mds_HN_tissue_6791[,1],mds_HN_tissue_6791[,2],bg=as.numeric(sex),pch=21,xlab="First dimension",ylab="Second dimension",cex=2,main="MultiDimension Scaling Plot: sex")
legend("bottomleft",levels(sex),col=seq(along=levels(sex)),pch=15,cex=0.7)

#anatomical location
plot(mds_HN_tissue_6791[,1],mds_HN_tissue_6791[,2],bg=as.numeric(tissue),pch=21,xlab="First dimension",ylab="Second dimension",cex=2,main="MultiDimension Scaling Plot: anatomical location")
legend("bottomleft",levels(tissue),col=seq(along=levels(tissue)),pch=15,cex=0.5)

# represent cancer and normal samples with different symbols and use colors for tissue
plot(mds_HN_tissue_6791[,1],mds_HN_tissue_6791[,2],col=as.numeric(tissue),pch=as.numeric(cancer_status_factors),xlab="First dimension",ylab="Second dimension",cex=3,main="MultiDimension Scaling Plot: anatomical location & cancer status")
legend("topleft",levels(tissue),col=seq(along=levels(tissue)),pch=15,cex=0.7)
legend("bottomleft",levels(cancer_status_factors),pch=seq(along=levels(cancer_status_factors)),cex=1)

hierarchial_cluster_tumor_normal <- hclust(dist_matrix_HN_tissue_6791)
plot(hierarchial_cluster_tumor_normal,cex=0.8,main="Hierarchical clustering of samples")
plot(hierarchial_cluster_tumor_normal,cex=0.8,main="Hierarchical clustering of samples",label=tissue)
```

**Exploratory Data Analysis for  HPVpositive vs HPVnegative**
```{r Chunk 5:  Exploratory Data Analysis for HPVpositive vs HPVnegative}
HNC_HPV_neg_gsmids <- rownames(subset(sample_info_HN_tissue,Case=="HN cancer" & HPV=="-",Case))
HNC_HPV_pos_gsmids <- rownames(subset(sample_info_HN_tissue,Case=="HN cancer" & HPV=="+",Case))

gset6791_HNCancer <- gset6791[,c(HNC_HPV_neg_gsmids,HNC_HPV_pos_gsmids)]
ex6791_HNCancer <- ex6791_HN_tissue[,c(HNC_HPV_neg_gsmids,HNC_HPV_pos_gsmids)]
sample_info_HNCancer <- sample_info_HN_tissue[c(HNC_HPV_neg_gsmids,HNC_HPV_pos_gsmids),]
setdiff(rownames(sample_info_HNCancer),colnames(ex6791_HNCancer))
HNCancer_HPV_status <- factor(sample_info_HNCancer$HPV, levels=c("-","+")) ## order of levels is important
HNCancer_tissue <- factor(sample_info_HNCancer$Anatomical.sites)
HPV_tissue <- factor(paste(HNCancer_tissue,HNCancer_HPV_status,sep=""))

dist_matrix_HNCancer <- dist(t(ex6791_HNCancer)) # very important to take transpose "t"
mds_HNCancer <- cmdscale(dist_matrix_HNCancer)

plot(mds_HNCancer[,1],mds_HNCancer[,2],bg=as.numeric(HNCancer_HPV_status),pch=21,xlab="First dimension",ylab="Second dimension",cex=2,main="MultiDimension Scaling Plot: HPV status")
legend("bottomleft",levels(HNCancer_HPV_status),col=seq(along=levels(HNCancer_HPV_status)),pch=15,cex=1)

plot(mds_HNCancer[,1],mds_HNCancer[,2],bg=as.numeric(HNCancer_tissue),pch=21,xlab="First dimension",ylab="Second dimension",cex=2,main="MultiDimension Scaling Plot: anatomical location")
legend("bottomright",levels(HNCancer_tissue),col=seq(along=levels(HNCancer_tissue)),pch=15,cex=0.8)

# represent hpv +ve and -ve samples with different symbols and use colors for tissue
plot(mds_HNCancer[,1],mds_HNCancer[,2],col=as.numeric(HNCancer_tissue),pch=as.numeric(HNCancer_HPV_status),xlab="First dimension",ylab="Second dimension",cex=3,main="MultiDimension Scaling Plot: anatomical location & HPV status")
legend("bottomright",levels(HNCancer_tissue),col=seq(along=levels(HNCancer_tissue)),pch=15,cex=0.7)
legend("bottomleft",levels(HNCancer_HPV_status),pch=seq(along=levels(HNCancer_HPV_status)),cex=1)
```

**Exploratory Data Analysis for  HPVpositive Tumor vs Normal (HPVnegative)**
```{r Chunk 6:  Exploratory Data Analysis for HPVpositive Tumor vs Normal (HPVnegative)}
gset6791_tumorHPVpositive_normal <- gset6791[,c(HNnormal_gsmids,HNC_HPV_pos_gsmids)]
sample_info_tumorHPVpositive_normal <- sample_info_HN_tissue[c(HNnormal_gsmids,HNC_HPV_pos_gsmids),]
cancer_status_tumorHPVpositive_normal <- factor(sample_info_tumorHPVpositive_normal$Case, levels=c("HN normal","HN cancer"))

ex6791_tumorHPVpositive_normal <- ex6791_HN_tissue[,c(HNnormal_gsmids,HNC_HPV_pos_gsmids)]
dist_matrix_tumorHPVpositive_normal <- dist(t(ex6791_tumorHPVpositive_normal)) # very important to take transpose "t"
mds_tumorHPVpositive_normal <- cmdscale(dist_matrix_tumorHPVpositive_normal)

#Cancer status
plot(mds_tumorHPVpositive_normal[,1],mds_tumorHPVpositive_normal[,2],bg=as.numeric(cancer_status_tumorHPVpositive_normal),pch=21,xlab="First dimension",ylab="Second dimension",cex=2,main="MultiDimension Scaling Plot: Cancer status")
legend("bottomleft",levels(cancer_status_tumorHPVpositive_normal),col=seq(along=levels(cancer_status_tumorHPVpositive_normal)),pch=15,cex=0.8)
# to identify which samples may not represent "pure normal" samples, add sample names to plot
text(mds_tumorHPVpositive_normal[,1],mds_tumorHPVpositive_normal[,2],labels=rownames(mds_tumorHPVpositive_normal),cex=0.5)
#identify(mds_tumorHPVpositive_normal)
sample_info_tumorHPVpositive_normal[c(28,4,29,2,3,17,27),c(2,4,6)]

# represent cancer and normal samples with different symbols and use colors for tissue. 
# See if GSM155718 is clustered with other cancer samples or is it clustered with other samples from same tissue.
plot(mds_tumorHPVpositive_normal[,1],mds_tumorHPVpositive_normal[,2],col=as.numeric(factor(sample_info_tumorHPVpositive_normal$Anatomical.sites)),pch=as.numeric(factor(sample_info_tumorHPVpositive_normal$Case)),xlab="First dimension",ylab="Second dimension",cex=3,main="MultiDimension Scaling Plot: Cancer & Tissue status")
legend("topleft",levels(factor(sample_info_tumorHPVpositive_normal$Anatomical.sites)),col=seq(along=levels(factor(sample_info_tumorHPVpositive_normal$Anatomical.sites))),pch=15,cex=0.7)
legend("bottomleft",levels(factor(sample_info_tumorHPVpositive_normal$Case)),pch=seq(along=levels(factor(sample_info_tumorHPVpositive_normal$Case))),cex=1)
```

**Exploratory Data Analysis for HPVnegative Tumor vs Normal (HPVnegative)**
```{r Chunk 7:  Exploratory Data Analysis for HPVnegative Tumor vs Normal (HPVnegative)}
gset6791_tumorHPVnegative_normal <- gset6791[,c(HNnormal_gsmids,HNC_HPV_neg_gsmids)]
sample_info_tumorHPVnegative_normal <- sample_info_HN_tissue[c(HNnormal_gsmids,HNC_HPV_neg_gsmids),]
cancer_status_tumorHPVnegative_normal <- factor(sample_info_tumorHPVnegative_normal$Case, levels=c("HN normal","HN cancer"))

ex6791_tumorHPVnegative_normal <- ex6791_HN_tissue[,c(HNnormal_gsmids,HNC_HPV_neg_gsmids)]
dist_matrix_tumorHPVnegative_normal <- dist(t(ex6791_tumorHPVnegative_normal)) # very important to take transpose "t"
mds_tumorHPVnegative_normal <- cmdscale(dist_matrix_tumorHPVnegative_normal)

# Cancer status
plot(mds_tumorHPVnegative_normal[,1],mds_tumorHPVnegative_normal[,2],bg=as.numeric(cancer_status_tumorHPVnegative_normal),pch=21,xlab="First dimension",ylab="Second dimension",cex=2,main="MultiDimension Scaling Plot: Cancer status")
legend("bottomleft",levels(cancer_status_tumorHPVnegative_normal),col=seq(along=levels(cancer_status_tumorHPVnegative_normal)),pch=15,cex=0.8)
# to identify which samples may not represent "pure normal" samples, add sample names to plot
text(mds_tumorHPVnegative_normal[,1],mds_tumorHPVnegative_normal[,2],labels=rownames(mds_tumorHPVnegative_normal),cex=0.6)
#identify(mds_tumorHPVnegative_normal)
sample_info_tumorHPVnegative_normal[c(3,18,40,15,38,4,17,16,2,15,19),c(2,4,6)]

# represent cancer and normal samples with different symbols and use colors for tissue. 
# See if GSM155718 & GSM155717 are clustered with other cancer samples or is it clustered with other samples from same tissue.
plot(mds_tumorHPVnegative_normal[,1],mds_tumorHPVnegative_normal[,2],col=as.numeric(factor(sample_info_tumorHPVnegative_normal$Anatomical.sites)),pch=as.numeric(factor(sample_info_tumorHPVnegative_normal$Case)),xlab="First dimension",ylab="Second dimension",cex=3,main="MultiDimension Scaling Plot: Cancer & Tissue status")
legend("topleft",levels(factor(sample_info_tumorHPVnegative_normal$Anatomical.sites)),col=seq(along=levels(factor(sample_info_tumorHPVnegative_normal$Anatomical.sites))),pch=15,cex=0.7)
legend("bottomleft",levels(factor(sample_info_tumorHPVnegative_normal$Case)),pch=seq(along=levels(factor(sample_info_tumorHPVnegative_normal$Case))),cex=0.8)
```

**Exploratory Data Analysis for Tumor vs Normal : Tonsil Tissue only**
```{r Chunk 8:  Exploratory Data Analysis for Tumor vs Normal : Tonsil Tissue only}
tonsil_normal_gsmids <-  rownames( subset(sample_info_HN_tissue,Anatomical.sites=="Tonsil" & Case=="HN normal",)  )
tonsil_cancer_gsmids <-  rownames( subset(sample_info_HN_tissue,Anatomical.sites=="Tonsil" & Case=="HN cancer",)  )
  
gset6791_tonsil_tumor_normal <- gset6791[,c(tonsil_normal_gsmids,tonsil_cancer_gsmids)]
sample_info_tonsil_tumor_normal <- subset(sample_info_HN_tissue,Anatomical.sites=="Tonsil")
cancer_status_tonsil_tumor_normal <- factor(sample_info_tonsil_tumor_normal$Case, levels=c("HN normal","HN cancer"))

ex6791_tonsil_tumor_normal <- ex6791_HN_tissue[,c(tonsil_normal_gsmids,tonsil_cancer_gsmids)]
dist_matrix_tonsil_tumor_normal <- dist(t(ex6791_tonsil_tumor_normal)) # very important to take transpose "t"
mds_tonsil_tumor_normal <- cmdscale(dist_matrix_tonsil_tumor_normal)

plot(mds_tonsil_tumor_normal[,1],mds_tonsil_tumor_normal[,2],bg=as.numeric(cancer_status_tonsil_tumor_normal),pch=21,xlab="First dimension",ylab="Second dimension",cex=2,main="MultiDimension Scaling Plot: Tonsil Samples, Cancer status")
legend("topleft",levels(cancer_status_tonsil_tumor_normal),col=seq(along=levels(cancer_status_tonsil_tumor_normal)),pch=15,cex=1)
text(mds_tonsil_tumor_normal[,1],mds_tonsil_tumor_normal[,2],labels=rownames(mds_tonsil_tumor_normal),cex=0.8)

# dendrogram to see which samples are close or far
hierarchial_cluster_tonsil_tumor_normal <- hclust(dist_matrix_tonsil_tumor_normal)
plot(hierarchial_cluster_tonsil_tumor_normal,cex=0.8,main="Hierarchical clustering of samples",label=cancer_status_tonsil_tumor_normal)
plot(hierarchial_cluster_tonsil_tumor_normal,cex=0.8,main="Hierarchical clustering of samples")
#sample_info_tonsil_tumor_normal[,c(2:9)]

tonsil_HPV_status <- factor(sample_info_tonsil_tumor_normal$HPV, levels=c("-","+"))
tonsil_cancer_HPV_status <- factor(paste(cancer_status_tonsil_tumor_normal,tonsil_HPV_status,sep=""),levels=c("HN normal-","HN cancer-","HN cancer+"))

plot(mds_tonsil_tumor_normal[,1],mds_tonsil_tumor_normal[,2],bg=as.numeric(tonsil_cancer_HPV_status),pch=21,xlab="First dimension",ylab="Second dimension",cex=2,main="MultiDimension Scaling Plot: Tonsil Samples, Cancer & HPV status")
legend("topleft",levels(tonsil_cancer_HPV_status),col=seq(along=levels(tonsil_cancer_HPV_status)),pch=15,cex=1)
text(mds_tonsil_tumor_normal[,1],mds_tonsil_tumor_normal[,2],labels=rownames(mds_tonsil_tumor_normal),cex=0.8)
plot(hierarchial_cluster_tonsil_tumor_normal,cex=0.8,main="Hierarchical clustering of samples",label=tonsil_cancer_HPV_status)

#tonsil_cancer_HPV_status
tonsil_cancer_HPVpositive_gsmids <-  rownames( subset(sample_info_HN_tissue,Anatomical.sites=="Tonsil" & Case=="HN cancer" & HPV=="+",)  )
tonsil_cancer_HPVnegative_gsmids <-  rownames( subset(sample_info_HN_tissue,Anatomical.sites=="Tonsil" & Case=="HN cancer" & HPV=="-",)  )
tonsil_cancer_HPVpositive_status <- factor(tonsil_cancer_HPV_status[c(1:4,8:10)],levels=c("HN normal-","HN cancer+"))
tonsil_cancer_HPVnegative_status <- factor(tonsil_cancer_HPV_status[c(1:4,5:7)],levels=c("HN normal-","HN cancer-"))
```

```{r Chunk 9: function to run limma}
RunLimma <- function(gset,groups){
  design_matrix <- model.matrix(~0+groups)
  colnames(design_matrix) <- c("CONTROL","TREATMENT")
  # to check order is correct
  # head(design_matrix)
  
  ## fit linear model for each gene given a series of arrays
  fit <- lmFit(gset, design_matrix)
  
  ## tell which levels should be compared
  colnames(design_matrix)
  cont.matrix <- makeContrasts(TREATMENT-CONTROL, levels=design_matrix)
  
  ## given a linear model fit to microarray data, compute estimated coefficients & standard errors for given set of contratsts
  fit2 <- contrasts.fit(fit, cont.matrix)
  fit2 <- eBayes(fit2, 0.01)
  
  ## to get the no.of genes in each array
  nrow_gset <- length(featureNames(gset))
  tT <- topTable(fit2, adjust="fdr", sort.by="logFC",number=nrow_gset)
  colnames(tT)
  tT <- subset(tT, select=c("ID","adj.P.Val","P.Value","logFC","t","B","AveExpr","ENTREZ_GENE_ID","Gene.Symbol","Gene.Title"))
  return(tT)
}

#function to remove rows which are not associated with any gene 
RemoveRows <- function(MatrixName,ColumnName){ #ColumnName is name of column containing gene id
  # remove rows that contain "NA" in any column
  complete <- complete.cases(MatrixName)
  Matrix_complete <- MatrixName[complete,]
  # remove rows that contain "" in column for gene id
  nonempty_gene_ids <- Matrix_complete[,ColumnName]!=""
  Matrix2 <- Matrix_complete[nonempty_gene_ids,]
  
  return(Matrix2)
}
```

**HPVpositive vs HPVnegative**
```{r Chunk 10: running limma for HPVpositive vs HPVnegative}
gse6791_HPV_positiveVsnegative <- RunLimma(gset6791_HNCancer,HNCancer_HPV_status) 
gse6791_HPV_positiveVsnegativeDEG <- RemoveRows(gse6791_HPV_positiveVsnegative,"ENTREZ_GENE_ID")
# volcano plot
plot(gse6791_HPV_positiveVsnegativeDEG$logFC,-log10(gse6791_HPV_positiveVsnegativeDEG$adj.P.Val), xlab="Effect size",ylab="- log (base 10) adj.p-values",main="HPVpositive vs HPVnegative")

gse6791_HPV_positiveVsnegativeDEGFinal <- subset(gse6791_HPV_positiveVsnegativeDEG,adj.P.Val<0.05 & abs(logFC)>0.585,)
length(unique(subset(gse6791_HPV_positiveVsnegativeDEGFinal,logFC>0.585,)$ENTREZ_GENE_ID))
length(unique(subset(gse6791_HPV_positiveVsnegativeDEGFinal,logFC<(-0.585),)$ENTREZ_GENE_ID))

# save(gse6791_HPV_positiveVsnegativeDEGFinal, file="./rough/HPVpositiveVsNegative/gse6791_HPV_positiveVsnegativeDEGFinal.rda")
# write.table(gse6791_HPV_positiveVsnegativeDEGFinal, file="./rough/gse6791_HPV_positiveVsnegativeDEGFinal.txt", row.names=F, sep="\t")

```

**Tumor vs Normal**
```{r Chunk 11: running limma for Tumor vs Normal }
gse6791_cancerVsnormal <- RunLimma(gset6791_HN_tissue,cancer_status_factors) 
gse6791_cancerVsnormalDEG <- RemoveRows(gse6791_cancerVsnormal,"ENTREZ_GENE_ID")
# volcano plot
plot(gse6791_cancerVsnormalDEG$logFC,-log10(gse6791_cancerVsnormalDEG$adj.P.Val), xlab="Effect size",ylab="- log (base 10) adj.p-values",main="Tumor vs Normal")

gse6791_cancerVsnormalDEGFinal <- subset(gse6791_cancerVsnormalDEG,adj.P.Val<0.05 & abs(logFC)>0.585,)
length(unique(subset(gse6791_cancerVsnormalDEGFinal,logFC>0.585,)$ENTREZ_GENE_ID))
length(unique(subset(gse6791_cancerVsnormalDEGFinal,logFC<(-0.585),)$ENTREZ_GENE_ID))
```

**HPVpositive Tumor vs Normal**
```{r Chunk 12: running limma for HPVpositive Tumor vs Normal (HPVnegative)}
gse6791_tumorHPVpositiveVsnormal <- RunLimma(gset6791_tumorHPVpositive_normal,cancer_status_tumorHPVpositive_normal) 
gse6791_tumorHPVpositiveVsnormalDEG <- RemoveRows(gse6791_tumorHPVpositiveVsnormal,"ENTREZ_GENE_ID")
# volcano plot
plot(gse6791_tumorHPVpositiveVsnormalDEG$logFC,-log10(gse6791_tumorHPVpositiveVsnormalDEG$adj.P.Val), xlab="Effect size",ylab="- log (base 10) adj.p-values",main="HPVpositive Tumor vs Normal")

gse6791_tumorHPVpositiveVsnormalDEGFinal <- subset(gse6791_tumorHPVpositiveVsnormalDEG,adj.P.Val<0.05 & abs(logFC)>0.585,)
length(unique(subset(gse6791_tumorHPVpositiveVsnormalDEGFinal,logFC>0.585,)$ENTREZ_GENE_ID))
length(unique(subset(gse6791_tumorHPVpositiveVsnormalDEGFinal,logFC<(-0.585),)$ENTREZ_GENE_ID))

# save(gse6791_tumorHPVpositiveVsnormalDEGFinal, file="./rough/tumorHPVpositiveVsnormal/gse6791_tumorHPVpositiveVsnormalDEGFinal.rda")
# write.table(gse6791_tumorHPVpositiveVsnormalDEGFinal, file="./rough/tumorHPVpositiveVsnormal/gse6791_tumorHPVpositiveVsnormalDEGFinal.txt", row.names=F, sep="\t")

```

**HPVnegative Tumor vs Normal**
```{r Chunk 13: running limma for HPVnegative Tumor vs Normal (HPVnegative)}
gse6791_tumorHPVnegativeVsnormal <- RunLimma(gset6791_tumorHPVnegative_normal,cancer_status_tumorHPVnegative_normal) 
gse6791_tumorHPVnegativeVsnormalDEG <- RemoveRows(gse6791_tumorHPVnegativeVsnormal,"ENTREZ_GENE_ID")
# volcano plot
plot(gse6791_tumorHPVnegativeVsnormalDEG$logFC,-log10(gse6791_tumorHPVnegativeVsnormalDEG$adj.P.Val), xlab="Effect size",ylab="- log (base 10) adj.p-values",main="HPVnegative Tumor vs Normal")

gse6791_tumorHPVnegativeVsnormalDEGFinal <- subset(gse6791_tumorHPVnegativeVsnormalDEG,adj.P.Val<0.05 & abs(logFC)>0.585,)
length(unique(subset(gse6791_tumorHPVnegativeVsnormalDEGFinal,logFC>0.585,)$ENTREZ_GENE_ID))
length(unique(subset(gse6791_tumorHPVnegativeVsnormalDEGFinal,logFC<(-0.585),)$ENTREZ_GENE_ID))

# save(gse6791_tumorHPVnegativeVsnormalDEGFinal, file="./rough/tumorHPVnegativeVsnormal/gse6791_tumorHPVnegativeVsnormalDEGFinal.rda")
# write.table(gse6791_tumorHPVnegativeVsnormalDEGFinal, file="./rough/tumorHPVnegativeVsnormal/gse6791_tumorHPVnegativeVsnormalDEGFinal.txt", row.names=F, sep="\t")
```

**Tumor vs Normal**
```{r Chunk 14: running limma for Tumor vs Normal for Tonsil Samples Only}
gse6791_tonsil_cancerVsnormal <- RunLimma(gset6791_tonsil_tumor_normal,cancer_status_tonsil_tumor_normal) 
gse6791_tonsil_cancerVsnormalDEG <- RemoveRows(gse6791_tonsil_cancerVsnormal,"ENTREZ_GENE_ID")
# volcano plot
plot(gse6791_tonsil_cancerVsnormalDEG$logFC,-log10(gse6791_tonsil_cancerVsnormalDEG$adj.P.Val), xlab="Effect size",ylab="- log (base 10) adj.p-values",main="Tumor vs Nnormal - Tonsil Samples")

gse6791_tonsil_cancerVsnormalDEGFinal <- subset(gse6791_tonsil_cancerVsnormalDEG,adj.P.Val<0.05 & logFC>0.585,)
length(unique(subset(gse6791_tonsil_cancerVsnormalDEGFinal,logFC>0.585,)$ENTREZ_GENE_ID))
length(unique(subset(gse6791_tonsil_cancerVsnormalDEGFinal,logFC<(-0.585),)$ENTREZ_GENE_ID))
```

**HeatMaps**
```{r Chunk15: Load packages and common commands for generating heatmaps. Try Heatplus package to add more annotation}
library(RColorBrewer)
library(gplots)
# color gradient to represent the expression values of gene 
hmcol <- colorRampPalette(brewer.pal(9,"GnBu"))(100)
```

```{r Chunk 16: Heat map for HPV+ve vs HPV-ve}
HPVpositiveVsnegative_DEG_ids <- rownames(gse6791_HPV_positiveVsnegativeDEGFinal)

# colors to represent different levels represented by columns - here: normal and cancer
HPVpositiveVsnegative_cols <- palette(brewer.pal(8,"Dark2"))[as.numeric(HNCancer_HPV_status)]
HNCancer_HPV_status
HPVpositiveVsnegative_cols

heatmap.2(ex6791_HNCancer[HPVpositiveVsnegative_DEG_ids,], labCol=HNCancer_HPV_status,trace="none",ColSideColors=HPVpositiveVsnegative_cols,col=hmcol,margin=c(6,6),cexCol=1.5)
# colors to represent different levels represented by columns - here: tissue and hpv status combined
HPVpositiveVsnegative_cols2 <- palette(brewer.pal(12,"Set3"))[as.numeric(HPV_tissue)]
heatmap.2(ex6791_HNCancer[HPVpositiveVsnegative_DEG_ids,], labCol=HPV_tissue,trace="none",ColSideColors=HPVpositiveVsnegative_cols2,col=hmcol,margin=c(10,6),cexCol=1.2)
```

```{r Chunk 17: Heat map for HPV+ve vs normal}
#DEG_ids <- gse6791_tumorHPVpositiveVsnormalDEGFinal$ID # Dnt do this, it gives wrong results
tumorHPVpositiveVsnormal_DEG_ids <- rownames(gse6791_tumorHPVpositiveVsnormalDEGFinal)
tumorHPVpositiveVsnormal_disease_status <- sample_info_tumorHPVpositive_normal$Case
tumorHPVpositiveVsnormal_disease_status <- gsub("HN ","",tumorHPVpositiveVsnormal_disease_status)
tumorHPVpositiveVsnormal_tissue_status <- sample_info_tumorHPVpositive_normal$Anatomical.sites
tumorHPVpositiveVsnormal_disease_tissue_status <- paste(tumorHPVpositiveVsnormal_disease_status,tumorHPVpositiveVsnormal_tissue_status,sep="-")
table(tumorHPVpositiveVsnormal_disease_tissue_status)

# colors to represent different levels represented by columns - here: normal and cancer
tumorHPVpositiveVsnormal_cols <- palette(brewer.pal(8,"Dark2"))[as.fumeric(tumorHPVpositiveVsnormal_disease_status)]
tumorHPVpositiveVsnormal_disease_status
tumorHPVpositiveVsnormal_cols
# dendrogram of samples using top 1000 DEG (Differentially Expressed Genes)
heatmap.2(ex6791_tumorHPVpositive_normal[tumorHPVpositiveVsnormal_DEG_ids,][1:1000,], labCol=tumorHPVpositiveVsnormal_disease_status,trace="none",ColSideColors=tumorHPVpositiveVsnormal_cols,col=hmcol,margin=c(6,6),cexCol=1.5)

# colors to represent different levels represented by columns - here: tissue and cancer status combined
tumorHPVpositiveVsnormal_cols2 <- palette(brewer.pal(12,"Set3"))[as.fumeric(tumorHPVpositiveVsnormal_disease_tissue_status)]
heatmap.2(ex6791_tumorHPVpositive_normal[tumorHPVpositiveVsnormal_DEG_ids,][1:1000,], labCol=tumorHPVpositiveVsnormal_disease_tissue_status,trace="none",ColSideColors=tumorHPVpositiveVsnormal_cols2,col=hmcol,margin=c(12,6),cexCol=1.2)
```

```{r chunk18: Heat map for HPVnegative vs normal}
tumorHPVnegativeVsnormal_DEG_ids <- rownames(gse6791_tumorHPVnegativeVsnormalDEGFinal)

tumorHPVnegativeVsnormal_disease_status <- sample_info_tumorHPVnegative_normal$Case
tumorHPVnegativeVsnormal_disease_status <- gsub("HN ","",tumorHPVnegativeVsnormal_disease_status)
tumorHPVnegativeVsnormal_tissue_status <- sample_info_tumorHPVnegative_normal$Anatomical.sites
tumorHPVnegativeVsnormal_disease_tissue_status <- paste(tumorHPVnegativeVsnormal_disease_status,tumorHPVnegativeVsnormal_tissue_status,sep="-")
table(tumorHPVnegativeVsnormal_disease_tissue_status)

# colors to represent different levels represented by columns - here: normal and cancer
tumorHPVnegativeVsnormal_cols <- palette(brewer.pal(8,"Dark2"))[as.fumeric(tumorHPVnegativeVsnormal_disease_status)]
tumorHPVnegativeVsnormal_cols
tumorHPVnegativeVsnormal_disease_status
heatmap.2(ex6791_tumorHPVnegative_normal[tumorHPVnegativeVsnormal_DEG_ids,], labCol=tumorHPVnegativeVsnormal_disease_status,trace="none",ColSideColors=tumorHPVnegativeVsnormal_cols,col=hmcol,margin=c(6,6),cexCol=1.5)

# colors to represent different levels represented by columns - here: tissue and cancer status combined
tumorHPVnegativeVsnormal_cols2 <- palette(brewer.pal(12,"Set3"))[as.fumeric(tumorHPVnegativeVsnormal_disease_tissue_status)]
heatmap.2(ex6791_tumorHPVnegative_normal[tumorHPVnegativeVsnormal_DEG_ids,], labCol=tumorHPVnegativeVsnormal_disease_tissue_status,trace="none",ColSideColors=tumorHPVnegativeVsnormal_cols2,col=hmcol,margin=c(12,6),cexCol=1.2)
```

